---
title: "Untitled"
output: html_document
---


```{r}
library(rethinking)
sppnames <- c( "afarensis","africanus","habilis","boisei",
               "rudolfensis","ergaster","sapiens")
brainvolcc <- c( 438 , 452 , 612, 521, 752, 871, 1350 )
masskg <- c( 37.0 , 35.5 , 34.5 , 41.5 , 55.5 , 61.0 , 53.5 )
d <- data.frame( species=sppnames , brain=brainvolcc , mass=masskg )
```

A standing question, however, is to what extent particular species have brains that
are larger than we’d expect, after taking body size into account. A common solution is to fit a linear regression that models brain size as a linear function of body size. Then the remaining variation in brain size can be modeled as a function of other variables, like ecology or diet. This is the same “statistical control” strategy explained in previous chapters.

The simplest model that relates brain size to body size is the linear one. It will be the
first model we consider. Before writing out the model, let’s rescale the variables. In this case, we want to standardize body mass—give it mean zero and standard deviation one—and rescale the outcome, brain volume, so that the largest observed value is 1. Why not standardize brain volume as well? Because we want to preserve zero as a reference point: No brain at all. You can’t have negative brain. I don’t think.

```{r}
d$mass_std <- (d$mass - mean(d$mass))/sd(d$mass)
d$brain_std <- d$brain / max(d$brain)
```
# Simple Linear Model
```{r}
flist <- alist(brain_std ~ dnorm(mu, sigma),
               mu <- alpha + b1 * mass_std,
               alpha ~ dnorm(.5, 1),
               b1 ~ dnorm(0, 10),
               sigma ~ dlnorm(0,1))

model <- quap(flist, data = d)


```

You could use R’s simple lm function to get the posterior distribution for m6.1. You won’t get a posterior for sigma however.
```{r}
m7.1_OLS <- lm( brain_std ~ mass_std , data=d )
post <- extract.samples( m7.1_OLS )
```

```{r}
R2_is_bad <- function( quap_fit ) {
s <- sim( quap_fit , refresh=0 )
r <- apply(s,2,mean) - d$brain_std
1 - var2(r)/var2(d$brain_std)
}
```

```{r}
m7.2 <- quap(
alist(
brain_std ~ dnorm( mu , exp(log_sigma) ),
mu <- a + b[1]*mass_std + b[2]*mass_std^2,
a ~ dnorm( 0.5 , 1 ),
b ~ dnorm( 0 , 10 ),
log_sigma ~ dnorm( 0 , 1 )
), data=d , start=list(b=rep(0,2)) )


m7.3 <- quap(
alist(
brain_std ~ dnorm( mu , exp(log_sigma) ),
mu <- a + b[1]*mass_std + b[2]*mass_std^2 +
b[3]*mass_std^3,
a ~ dnorm( 0.5 , 1 ),
b ~ dnorm( 0 , 10 ),
log_sigma ~ dnorm( 0 , 1 )
), data=d , start=list(b=rep(0,3)) )

m7.4 <- quap(
alist(
brain_std ~ dnorm( mu , exp(log_sigma) ),
mu <- a + b[1]*mass_std + b[2]*mass_std^2 +
b[3]*mass_std^3 + b[4]*mass_std^4,
a ~ dnorm( 0.5 , 1 ),
b ~ dnorm( 0 , 10 ),
log_sigma ~ dnorm( 0 , 1 )
), data=d , start=list(b=rep(0,4)) )

m7.5 <- quap(
alist(
brain_std ~ dnorm( mu , exp(log_sigma) ),
mu <- a + b[1]*mass_std + b[2]*mass_std^2 +
b[3]*mass_std^3 + b[4]*mass_std^4 +
b[5]*mass_std^5,
a ~ dnorm( 0.5 , 1 ),
b ~ dnorm( 0 , 10 ),
log_sigma ~ dnorm( 0 , 1 )
), data=d , start=list(b=rep(0,5)) )

m7.6 <- quap(
alist(
brain_std ~ dnorm( mu , 0.001 ),
mu <- a + b[1]*mass_std + b[2]*mass_std^2 +
b[3]*mass_std^3 + b[4]*mass_std^4 +
b[5]*mass_std^5 + b[6]*mass_std^6,
a ~ dnorm( 0.5 , 1 ),
b ~ dnorm( 0 , 10 )
), data=d , start=list(b=rep(0,6)) )
```

Now to plot each model.
```{r}
post <- extract.samples(m7.6)
mass_seq <- seq( from=min(d$mass_std) , to=max(d$mass_std) , length.out=100 )
l <- link( m7.6 , data=list( mass_std=mass_seq ) )
mu <- apply( l , 2 , mean )
ci <- apply( l , 2 , PI )
plot( brain_std ~ mass_std , data=d )
lines( mass_seq , mu )
shade( ci , mass_seq )
```
