---
title: "Chaper 12"
output: pdf_document
---

# 1. Overdispersion
```{r}
library(rethinking)
pbar <- .5
theta <- 5
curve( dbeta2(x,pbar,theta) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
```


(None of the overdispersion models in the chapter will run for me...)

Recall that in the previous chapter, a binomial model of these data that omitted department ended up misleading, because there is an indirect path from gender through department to admission. That confound resulted in a spurious indication that female applicants had lower odds of admission. But the model above is not confounded, despite not containing the department variable. How is this?


# 2. Zero-Inflated Outcomes

```{r}
# define parameters
prob_drink <- 0.2 # 20% of days
rate_work <- 1 # average 1 manuscript per day
# sample one year of production
N <- 365
# simulate days monks drink
set.seed(365)
drink <- rbinom( N , 1 , prob_drink )
# simulate manuscripts completed
y <- (1-drink)*rpois( N , rate_work )
```


```{r}
library(rethinking)
m12.3 <- ulam(
alist(
y ~ dzipois( p , lambda ),
logit(p) <- ap,
log(lambda) <- al,
ap ~ dnorm( -1.5 , 1 ),
al ~ dnorm( 1 , 0.5 )
) , data=list(y=y) , chains=4 )
```





```{r}
precis(m12.3)
```



We can recover the parameters used to generate the sample data (20% chance of drinking, 1 manuscript per day)
```{r}
post <- extract.samples( m12.3 )
mean( inv_logit( post$ap ) ) # probability drink
mean( exp( post$al ) ) # rate finish manuscripts, when not drinking
```

# 3 Ordered Categorical Outcomes

The trolly problem.

```{r}
data(Trolley)
d <- Trolley
head(d)
```



```{r}
simplehist( d$response , xlim=c(1,7) , xlab="response" )
```

```{r}
# discrete proportion of each response value
pr_k <- table( d$response ) / nrow(d)
# cumsum converts to cumulative proportions
cum_pr_k <- cumsum( pr_k )
# plot
plot( 1:7 , cum_pr_k , type="b" , xlab="response" ,
ylab="cumulative proportion" , ylim=c(0,1) )
```

```{r}
logit <- function(x) log(x/(1-x)) # convenience function
plot(round( lco <- logit( cum_pr_k ) , 2 ), type = 'b')
```

Beta-binomial and gamma-poisson (negative binomial) are ways to cope with over-dispersion.








